Compositional equivalences based on Open pNets\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-04-09 2016-Weak-FH-Bisim/WeakBisim.tex
biblio, TODOs, removed \I et def 11 corrigee

\paragraph{Previous Works and Contribution}
While most of our previous works relied on closed, fully-instantiated semantics~\cite{BBCHM:article2009,AmeurBoulifa2017,HKM-FASE16}, it is only recently that we could design a first version of a  parameterised semantics for pNets with a strong bisimulation equivalence~\cite{henrio:Forte2016}. This article builds open this previous parameterised semantics and provide a cleaner version of the semantics with a slightly simplified formalism and extends it to deal with weak bisimulation. The slight simplification of the formula allowed us to adopt a much cleaner presentation where the theory for open automata, the algebra used to define the semantics of pNets with holes, can be studied independently from its application to pNets. Also, in previous work the study of compositionability was only partial, and in particular the proof that bisimulation is an equivalence is one new contribution of this article and provides a particularly interesting insight on the semantic model we use.
The new formalism allowed us to extend the work and define weak bisimulation  for open automata, which is entirely new. This allows us to define a weak bisimulation equivalence for open pNets with valuable properties of compositionality. 
To summarise, the contribution of this paper are the following
\begin{itemize}
\item The definition of open automata: an algebra of parameterised automata with hole with a strong bisimulation equivalence. This is an adaptation of~\cite{henrio:Forte2016} with an additional property stating that strong bisimulation equivalence is indeed an equivalence relation.
\item A semantics for open pNets expressed as translation to open automata. This is an adaptation of~\cite{henrio:Forte2016} with a complete proof that strong bisimulation is compositional.
\item A theory of weak bisimulation for open automata, and its properties. It relies on the definition of weak open transitions that are derived from transitions of the open automaton by concatenating invisible action transitions with one (visible or not) transition. The precise definition of the concatenation is also a major contribution of this article.
\item A resulting weak bisimulation equivalence for open pNets and a simple static condition on synchronisation vectors inside pNets that is sufficient to ensure that weak bisimulation is compositional.
\item An illustrative example \TODO{TODO}
\end{itemize}

@@

%\def\Talg{\mathcal{T}_{\Sigma}}
%\def\BoolExprs{\mathcal{B}}
%\def\Exprs{\mathcal{E}} \TODO{use Exprs}
%\def\Actionalg{\mathcal{A}}
\TODO{Attention: la notation of term algebra is the same for set of transition.}

@@

\begin{definition}[pNets]\label{def-pnets}
A pNet $\pNet$ is a hierarchical structure where leaves are pLTSs and holes:\\
$\pNet,\pNetQ\triangleq \pLTS~|~\mylangle \pNet_i^{i\in I}, \Sort_j^{j\in J}, \symb{SV}_k^{k\in 
K}\myrangle$
where
\begin{itemize}
\item[$\bullet$] $\pNet_i^{i\in I}$ is the family of sub-pNets indexed over $I \in \I$.
%  $\pNet_i^{i\in I}$ is a family of sub-pNets where $I\in\I_\P$ is the set over which sub-pNets are indexed.

\item[$\bullet$] $J\!\in\!\I_\P$ is a set of indexes, called \emph{holes}.
$I$ and $J$ are \emph{disjoint}: $I\!\cap\! J=\emptyset$,  $I\!\cup\! J\neq\emptyset$
\item[$\bullet$] $\Sort_j \subseteq \AlgAS$ is a set of action terms, denoting the 
\emph{sort} of
hole $j$.

\item[$\bullet$] $\symb{SV}_k^{k\in K}$ is a set of
  synchronisation vectors ($K\in\I_\P$). $\forall k\!\in\! K,
  \symb{SV}_k\!=\!\SV{\alpha_{l}^{l\in I_k \uplus J_k}}{\alpha'_k}{e_k}$ where
  $\alpha'_k\in \mathcal{A}_\P$, $I_k\subseteq I$, $J_k\subseteq J$,
  $\forall i\!\in\!
  I_k.\,\alpha_{i}\!\in\!\Sort(\pNet_i)$,  $\forall j\!\in\!
  J_k.\,\alpha_{j}\!\in\!\Sort_j$, and $\vars(\alpha'_k)\subseteq \bigcup_{l\in I_k\uplus 
  J_k}{\vars({\alpha_l})}$. The global action of a vector $\symb{SV}_k$ is
$\alpha'_k$. $e_k \in \AlgB$ is a guard associated to the vector such that
$\vars(e_k)\subseteq \bigcup_{l\in I_k\uplus J_k}{\vars({\alpha_l})}$.
\end{itemize}
Synchronisation vectors are identified modulo renaming of variables that appear in their 
action terms.
\end{definition}

@@

\subsection{Computing and using open automata}
\TODO{REMOVE?  published at Avocs, we can shorten and add a bibref}
In this section we present a simple algorithm to construct the open
automaton representing the behaviour of an open pNet, and we prove that
under reasonable conditions this automaton is finite.

\begin{algorithm}[Behavioural semantics of open pNets: Sketch]
This is a standard residual algorithm over a set of open-automaton
states, but where transitions are open transitions
constructively ``proven'' by deduction trees.

1) Start with a set of unexplored states containing the initial state
of the automaton, and an empty set of explored states.

2) While there are unexplored states, pick one state from the
unexplored set and add it to the explored set.

2a- From this state
build all possible deduction trees by application of the structural
rules \TrUn\ and \TrDeux, using all applicable combinations
of synchronisation vectors.
In practice, the algorithm does not explicitely construct the
deduction tree, but for each pNet node, builds all possible
combinations of the OTs of its subnets, and filter these combinations
by matching with all synchronisation vectors.

2b- For each OT built, add the transition in the
outgoing transitions of the current state, and add the
resulting state in the unexplored set if it is not already there.

3) The predicate is submitted to Z3 for checking satisfiability. If it
is NOT satisfiable, the resulting OT is discarded. This will minimize
the number of resulting transitions, and by reachability analysis,
reduce the resulting open automaton.

\end{algorithm}

Given an open-pNet
with finite synchronisation sets, finitely many leaves and
holes, and each pLTS at leaves having a finite number of states and
(symbolic) transitions, has a finite automaton (see \ref{henrio:Forte2016}).

@@

Finally, the previous theorems can be composed to state a general theorem about 
composability and FH-bisimilarity.
\begin{theorem}[Composability] \label{thm-composability}
	Consider two FH-bisimilar pNets with an arbitrary number of holes, when replacing, 
	inside those two original pNets, a subset of the holes by FH-bisimilar pNets, we 
	obtain two FH-bisimilar pNets.
\end{theorem}
This theorem is quite powerful somehow states that the theory of open pNets is convenient to study properties of process composition. Open pNets can indeed be used to study process operators and process algebras, as shown in~\cite{henrio:Forte2016}, or to study interaction protocols \TODO{ref vers fault tolerance byzantine FACS 2011}

@@

Intuition: tau is non observable if the automaton allows any tau transition from holes, and the global transition resulting from a tau in a hole is a tau transition not changing the action state.
We define $\Id(s)$ as the identity function on variables of state s.
\begin{definition}[Non-observability of silent actions]\label{def:Non-ObsTau}
An open automaton $A = <J,\mathcal{S},s_0,\mathcal{T}>$ \emph{cannot observe silent actions} if and only if for all $j$ in $J$ and $s$ in $\mathcal{S}$ we have
\begin{enumerate}
\item
\[ \openrule
         {
           (j\mapsto\tau),\True,\Id(s)}
         {s \OTarrow {\tau} s}
         \in \mathcal{T}
\]
and 
\item for all $\alpha$, $j$, $s$, $\Post$ and $s'$ such that
\[ \openrule
         {
           (j\mapsto\tau),\Pred,\Post}
         {s \OTarrow {\alpha} s'}
         \in \mathcal{T} \] we have \[ \alpha=\tau\land s=s'\land \Pred=\True\land\Post=\Id(s)
\]
\TODO{question : it looks like we can have 2 holes doing a tau and observe this ; I belive this is not safe TODO LUDO add a $A(j)=\tau$}
\end{enumerate}
\end{definition}
The first statement of the definition state that the open automaton must allow a hole to do a silent action at any time, and must not observe it, i.e. cannot change its internal state because a hole did a $\tau$ transition. The second statement ensures that there cannot be in the open automaton other transitions that would be able to observe a silent action from a hole.

@@

\begin{property}[pNets that cannot observe silent actions]
The open semantics of a pNets that cannot observe silent actions is an open automaton that  cannot observe silent actions.
\end{property}

\TODO{question : we could merge prop 1 and def above, is it better?}

>>>>

\paragraph{Previous Works and Contribution}
While most of our previous works relied on closed, fully-instantiated semantics~\cite{BBCHM:article2008,AmeurBoulifa2017,HKM-FASE16}, it is only recently that we could design a first version of a  parameterised semantics for pNets with a strong bisimulation equivalence~\cite{henrio:Forte2016}. This article builds open this previous parameterised semantics and provide a cleaner version of the semantics with a slightly simplified formalism and extends it to deal with weak bisimulation. The slight simplification of the formula allowed us to adopt a much cleaner presentation where the theory for open automata, the algebra used to define the semantics of pNets with holes, can be studied independently from its application to pNets. Also, in previous work the study of compositionability was only partial, and in particular the proof that bisimulation is an equivalence is one new contribution of this article and provides a particularly interesting insight on the semantic model we use.
The new formalism allowed us to extend the work and define weak bisimulation  for open automata, which is entirely new. This allows us to define a weak bisimulation equivalence for open pNets with valuable properties of compositionality. 
To summarise, the contribution of this paper are the following
\begin{itemize}
\item The definition of open automata: an algebra of parameterised automata with hole with a strong bisimulation equivalence. This is an adaptation of~\cite{henrio:Forte2016} with an additional property stating that strong bisimulation equivalence is indeed an equivalence relation.
\item A semantics for open pNets expressed as translation to open automata. This is an adaptation of~\cite{henrio:Forte2016} with a complete proof that strong bisimulation is compositional.
\item A theory of weak bisimulation for open automata, and its properties. It relies on the definition of weak open transitions that are derived from transitions of the open automaton by concatenating invisible action transitions with one (visible or not) transition. The precise definition of the concatenation is also a major contribution of this article.
\item A resulting weak bisimulation equivalence for open pNets and a simple static condition on synchronisation vectors inside pNets that is sufficient to ensure that weak bisimulation is compositional.
\item An illustrative example \TODO{TODO}
\end{itemize}

@@

%\def\Talg{\mathcal{T}_{\Sigma}}
%\def\BoolExprs{\mathcal{B}}
%\def\Exprs{\mathcal{E}} \TODO{use Exprs}
%\def\Actionalg{\mathcal{A}}

@@

\begin{definition}[pNets]\label{def-pnets}
A pNet $\pNet$ is a hierarchical structure where leaves are pLTSs and holes:\\
$\pNet,\pNetQ\triangleq \pLTS~|~\mylangle \pNet_i^{i\in I}, \Sort_j^{j\in J}, \symb{SV}_k^{k\in 
K}\myrangle$
where
\begin{itemize}
\item[$\bullet$] $\pNet_i^{i\in I}$ is the family of sub-pNets indexed over $I$.
%  $\pNet_i^{i\in I}$ is a family of sub-pNets where $I\in\I_\P$ is the set over which sub-pNets are indexed.

\item[$\bullet$] $J$ is a set of indexes, called \emph{holes}.
$I$ and $J$ are \emph{disjoint}: $I\!\cap\! J=\emptyset$,  $I\!\cup\! J\neq\emptyset$
\item[$\bullet$] $\Sort_j \subseteq \AlgAS$ is a set of action terms, denoting the 
\emph{sort} of
hole $j$.

\item[$\bullet$] $\symb{SV}_k^{k\in K}$ is a set of
  synchronisation vectors. $\forall k\!\in\! K,
  \symb{SV}_k\!=\!\SV{\alpha_{l}^{l\in I_k \uplus J_k}}{\alpha'_k}{e_k}$ where
  $\alpha'_k\in \mathcal{A}_\P$, $I_k\subseteq I$, $J_k\subseteq J$,
  $\forall i\!\in\!
  I_k.\,\alpha_{i}\!\in\!\Sort(\pNet_i)$,  $\forall j\!\in\!
  J_k.\,\alpha_{j}\!\in\!\Sort_j$, and $\vars(\alpha'_k)\subseteq \bigcup_{l\in I_k\uplus 
  J_k}{\vars({\alpha_l})}$. The global action of a vector $\symb{SV}_k$ is
$\alpha'_k$. $e_k \in \AlgB$ is a guard associated to the vector such that
$\vars(e_k)\subseteq \bigcup_{l\in I_k\uplus J_k}{\vars({\alpha_l})}$.
\end{itemize}
Synchronisation vectors are identified modulo renaming of variables that appear in their 
action terms.
\end{definition}

@@

%\subsection{Computing and using open automata}
%\TODO{REMOVE?  published at Avocs, we can shorten and add a bibref}
%In this section we present a simple algorithm to construct the open
%automaton representing the behaviour of an open pNet, and we prove that
%under reasonable conditions this automaton is finite.
%
%\begin{algorithm}[Behavioural semantics of open pNets: Sketch]
%This is a standard residual algorithm over a set of open-automaton
%states, but where transitions are open transitions
%constructively ``proven'' by deduction trees.
%
%1) Start with a set of unexplored states containing the initial state
%of the automaton, and an empty set of explored states.
%
%2) While there are unexplored states, pick one state from the
%unexplored set and add it to the explored set.
%
%2a- From this state
%build all possible deduction trees by application of the structural
%rules \TrUn\ and \TrDeux, using all applicable combinations
%of synchronisation vectors.
%In practice, the algorithm does not explicitely construct the
%deduction tree, but for each pNet node, builds all possible
%combinations of the OTs of its subnets, and filter these combinations
%by matching with all synchronisation vectors. 
% 
%2b- For each OT built, add the transition in the
%outgoing transitions of the current state, and add the
%resulting state in the unexplored set if it is not already there.
%
%3) The predicate is submitted to Z3 for checking satisfiability. If it
%is NOT satisfiable, the resulting OT is discarded. This will minimize
%the number of resulting transitions, and by reachability analysis,
%reduce the resulting open automaton.
%
%\end{algorithm}

Given an open-pNet
with finite synchronisation sets, finitely many leaves and
holes, and each pLTS at leaves having a finite number of states and
(symbolic) transitions, has a finite automaton (see \cite{henrio:Forte2016}). The algorithm for building such an automaton can be found in~\cite{QBMZ-AVOCS18}.

@@

Finally, the previous theorems can be composed to state a general theorem about 
composability and FH-bisimilarity.
\begin{theorem}[Composability] \label{thm-composability}
	Consider two FH-bisimilar pNets with an arbitrary number of holes, when replacing, 
	inside those two original pNets, a subset of the holes by FH-bisimilar pNets, we 
	obtain two FH-bisimilar pNets.
\end{theorem}
This theorem is quite powerful somehow states that the theory of open pNets is convenient to study properties of process composition. Open pNets can indeed be used to study process operators and process algebras, as shown in~\cite{henrio:Forte2016}, or to study interaction protocols~\cite{BHHM:FACS11}.

@@

Intuition: tau is non observable if the automaton allows any tau transition from holes, and the global transition resulting from a tau in a hole is a tau transition not changing the action state.
We define $\Id(s)$ as the identity function on variables of state s.
\begin{definition}[Non-observability of silent actions]\label{def:Non-ObsTau}
An open automaton $A = <J,\mathcal{S},s_0,\mathcal{T}>$ \emph{cannot observe silent actions} if and only if for all $j$ in $J$ and $s$ in $\mathcal{S}$ we have
\begin{enumerate}
\item
\[ \openrule
         {
           (j\mapsto\tau),\True,\Id(s)}
         {s \OTarrow {\tau} s}
         \in \mathcal{T}
\]
and 
\item for all $A$, $\alpha$,  $s$, $\Post$ and $s'$ such that
\[ \openrule
         {
           A,\Pred,\Post}
         {s \OTarrow {\alpha} s'}
         \in \mathcal{T} \] If there is a $j$ such that $A(j)=\tau$ then we have \[ \alpha=\tau\land s=s'\land \Pred=\True\land\Post=\Id(s)
\]
\end{enumerate}
\end{definition}
The first statement of the definition state that the open automaton must allow a hole to do a silent action at any time, and must not observe it, i.e. cannot change its internal state because a hole did a $\tau$ transition. The second statement ensures that there cannot be in the open automaton other transitions that would be able to observe a silent action from a hole.

@@

\begin{property}[Non-observability of silent actions]
The open semantics of a pNets that cannot observe silent actions is an open automaton that  cannot observe silent actions.
\end{property}

%\TODO{question : we could merge prop 1 and def above, is it better?}
